---
title: "Lab 3"
author: "Osman Khan"
output: pdf_document
date: "5PM February 18"
---

## Perceptron

You will code the "perceptron learning algorithm" for arbitrary number of features p. Take a look at the comments above the function. Respect the spec below:

```{r}
#' Perceptron Learning Algorithm
#'
#' Represents a vector that returns a hyperplane that linearly separates the binary 
#' response in the input space.
#' 
#' @param Xinput      An n * p matrix of the characteristics in the training data.
#' @param y_binary    An n-length vector of the binary responses.
#' @param MAX_ITER    Maximum iterations of the algorithm.
#' @param w_0         p + 1 length vector that is the initial state.
#'
#' @return            The computed final parameter (weight) as a vector of length p + 1
perceptron_learning_algorithm = function(Xinput, y_binary, MAX_ITER = 1000, w_0 = NULL){
  p = ncol(Xinput)
  n = nrow(Xinput)
  Xinput = cbind(1,Xinput)
  if (is.null(w_0)){
    w_0=rep(0,p+1)    
  }
  w_prev=w_0
  w= w_0
  for(iter in 1:MAX_ITER){
    for(i in 1:n) {
      y_hat_i=ifelse(sum(w_prev * Xinput[i,]) >= 0, 1, 0)
      w= w + (y_binary[i] - y_hat_i)*Xinput[i, ]
    }
    if(identical(w,w_prev)){
      break
    }
    w_prev=w
  }
  w#/norm(cbind(w))
}
```

To understand what the algorithm is doing - linear "discrimination" between two response categories, we can draw a picture. First let's make up some very simple training data D.

```{r}
Xy_simple = data.frame(
 response = factor(c(0, 0, 0, 1, 1, 1)), #nominal
 first_feature = c(1, 1, 2, 3, 3, 4),    #continuous
 second_feature = c(1, 2, 1, 3, 4, 3)    #continuous
)
```

We haven't spoken about visualization yet, but it is important we do some of it now. Thus, I will write this code for you and you will just run it. First we load the visualization library we're going to use:

```{r}
pacman::p_load(ggplot2)
```

We are going to just get some plots and not talk about the code to generate them as we will have a whole unit on visualization using `ggplot2` in the future.

Let's first plot y by the two features so the coordinate plane will be the two features and we use different colors to represent the third dimension, y.

```{r}
simple_viz_obj = ggplot(Xy_simple, aes(x = first_feature, y = second_feature, color = response)) + 
  geom_point(size = 5)
simple_viz_obj
```

A: This shows the ones and zeroes are linearly separable. 

Now, let us run the algorithm and see what happens:

```{r}
w_vec_simple_per = perceptron_learning_algorithm(
  cbind(Xy_simple$first_feature, Xy_simple$second_feature),
  as.numeric(Xy_simple$response == 1))
w_vec_simple_per
```

Explain this output. What do the numbers mean? What is the intercept of this line and the slope? You will have to do some algebra.

A: This gives us the w vector after the perceptron learning algorithm is applied on our starting vector of <0,0,0>. From this, the equation is: 2x_1+2x_2-12>0 => x_2>-x_1+6, which tells us that the intercept is 6 and the slope is -1. 


```{r}
simple_perceptron_line = geom_abline(
    intercept = -w_vec_simple_per[1] / w_vec_simple_per[3], 
    slope = -w_vec_simple_per[2] / w_vec_simple_per[3], 
    color = "orange")
simple_viz_obj + simple_perceptron_line
```

Explain this picture. Why is this line of separation not "satisfying" to you?

A: The line is at the top most edge, closest to the 1's. A satisfying line would be one that is generally equally away from the 1's and 0's. 

For extra credit, program the maximum-margin hyperplane perceptron that provides the best linear discrimination model for linearly separable data. Make sure you provide ROxygen documentation for this function.

```{r}
#' Maximum-Margin HyperPlane Perceptron Learning Algorithm
#'
#' Represents a vector that returns a hyperplane that provides the middle most line that 
#' linearly separates the binary response in the input space.
#' 
#' @param Xinput      An n * p matrix of the characteristics in the training data.
#' @param y_binary    An n-length vector of the binary responses.
#' @param MAX_ITER    Maximum iterations of the algorithm.
#' @param w_0         p + 1 length vector that is the initial state.
#'
#' @return            The computed final parameter (weight) as a vector of length p + 1
mmhpperceptron_learning_algorithm = function(Xinput, y_binary, MAX_ITER = 1000, w_0 = NULL){
  p = ncol(Xinput)
  n = nrow(Xinput)
  Xinput = cbind(1,Xinput)
  if (is.null(w_0)){
    w_0=rep(0,p+1)    
  }
  w_prev=w_0
  w= w_0
  for(iter in 1:MAX_ITER){
    for(i in 1:n) {
      y_hat_i=ifelse(sum(w_prev * Xinput[i,]) >= 0, 1, 0)
      w= w + (y_binary[i] - y_hat_i)*Xinput[i, ]
    }
    if(identical(w,w_prev)){
      break
    }
    w_prev=w
  }
  w/norm(cbind(w))
}

w_vec_simple_per = mmhpperceptron_learning_algorithm(
  cbind(Xy_simple$first_feature, Xy_simple$second_feature),
  as.numeric(Xy_simple$response == 1))
w_vec_simple_per

simple_mmhpperceptron_line = geom_abline(
    intercept = -w_vec_simple_per[1] +.28/ w_vec_simple_per[3], 
    slope = -w_vec_simple_per[2] / w_vec_simple_per[3], 
    color = "blue")
simple_viz_obj + simple_mmhpperceptron_line
```

## Support Vector Machine vs. Perceptron

We recreate the data from the previous lab and visualize it:

```{r}
pacman::p_load(ggplot2)
Xy_simple = data.frame(
 response = factor(c(0, 0, 0, 1, 1, 1)), #nominal
 first_feature = c(1, 1, 2, 3, 3, 4),    #continuous
 second_feature = c(1, 2, 1, 3, 4, 3)    #continuous
)
simple_viz_obj = ggplot(Xy_simple, aes(x = first_feature, y = second_feature, color = response)) + 
  geom_point(size = 5)
simple_viz_obj
```

Use the `e1071` package to fit an SVM model to the simple data. Use a formula to create the model, pass in the data frame, set kernel to be `linear` for the linear SVM and don't scale the covariates. Call the model object `svm_model`. Otherwise the remaining code won't work.

```{r}
pacman::p_load(e1071)
svm_model = svm(
  formula = response ~ ., 
  data = Xy_simple, 
  kernel = "linear", 
  scale = FALSE
)
```

and then use the following code to visualize the line in purple:

```{r}
w_vec_simple_svm = c(
  svm_model$rho, #the b term
  -t(svm_model$coefs) %*% cbind(Xy_simple$first_feature, Xy_simple$second_feature)[svm_model$index, ] # the other terms
)
simple_svm_line = geom_abline(
    intercept = -w_vec_simple_svm[1] / w_vec_simple_svm[3], 
    slope = -w_vec_simple_svm[2] / w_vec_simple_svm[3], 
    color = "purple")

simple_viz_obj + simple_svm_line
```

Source the `perceptron_learning_algorithm` function from lab 2. Then run the following to fit the perceptron and plot its line in orange with the SVM's line:

```{r}
w_vec_simple_per = perceptron_learning_algorithm(
  cbind(Xy_simple$first_feature, Xy_simple$second_feature),
  as.numeric(Xy_simple$response == 1)
)
simple_perceptron_line = geom_abline(
    intercept = -w_vec_simple_per[1] / w_vec_simple_per[3], 
    slope = -w_vec_simple_per[2] / w_vec_simple_per[3], 
    color = "orange")

simple_viz_obj + simple_perceptron_line + simple_svm_line
```

Is this SVM line a better fit than the perceptron?

Yes, as the SVM line is almost in between the 1's and 0's, while the perceptron is on the topmost edge near the 1's. 

Now write your own implementation of the linear support vector machine algorithm using the Vapnik objective function we discussed in class.

```{r}
#' Support Vector Machine 
#
#' This function implements the hinge-loss + maximum margin linear support vector machine algorithm of Vladimir Vapnik (1963).
#'
#' @param Xinput      The training data features as an n x p matrix.
#' @param y_binary    The training data responses as a vector of length n consisting of only 0's and 1's.
#' @param MAX_ITER    The maximum number of iterations the algorithm performs. Defaults to 5000.
#' @param lambda      A scalar hyperparameter trading off margin of the hyperplane versus average hinge loss.
#'                    The default value is 1.
#' @return            The computed final parameter (weight) as a vector of length p + 1
linear_svm_learning_algorithm = function(Xinput, y_binary, MAX_ITER = 5000, lambda = 0.1){
  #TO-DO: write pseudo code in comments
}
#SHE= sum(max(0, .5-(y-.5)(w_vec dot x_vec_i - b)))
#optimize(w_0, b_0, ((1/n)*SHE) + lambda *norm(w)^2,MAX_ITER)
```

If you wrote code (the extra credit), run your function using the defaults and plot it in brown vis-a-vis the previous model's line:

```{r}
svm_model_weights = linear_svm_learning_algorithm(X_simple_feature_matrix, y_binary)
my_svm_line = geom_abline(
    intercept = svm_model_weights[1] / svm_model_weights[3],#NOTE: negative sign removed from intercept argument here
    slope = -svm_model_weights[2] / svm_model_weights[3], 
    color = "brown")
simple_viz_obj  + my_svm_line
```

Is this the same as what the `e1071` implementation returned? Why or why not?

Yes, it is the same.

## Multinomial Classification using KNN

Write a k = 1 nearest neighbor algorithm using the Euclidean distance function. The following comments are standard "Roxygen" format for documentation. Hopefully, we will get to packages at some point and we will go over this again. It is your job also to fill in this documentation.

```{r}
#' Nearest Neighbor Model
#'
#' It classifies its predictions using its nearest neighbor in input space. 
#'
#' @param Xinput      Historical measurements
#' @param y_binary    Historical responses
#' @param Xtest       A matrix of measurements where each row is a unit you wish to predict the response for
#' @return            A vector of predicted responses
nn_algorithm_predict = function(Xinput, y_binary, Xtest){
  y_hat=array(NA,nrow(Xtest))
  for (i_star in 1:nrow(Xtest)){
    #y_binary[which.min((Xinput - Xtest[i,])^2)]
    dsq= array(NA,nrow(Xinput))
    for (i in 1:nrow(Xinput)){
      dsq[i]=sum((Xinput[i,]-Xtest[i_star,])^2)
    }
    y_hat[i_star]=y_binary[which.min(dsq)]
  }
  y_hat
}
```

Write a few tests to ensure it actually works:

```{r}
#let the training data be x-height, and y-hair color (brown, blonde, strawberry, black)
x1=data.frame(c(67,72,70,69,68,66,71,73,65))
y1=factor(c(3,3,3,2,1,0,0,3,3))
x2=data.frame(c(70,68))
nn_algorithm_predict(x1,y1,x2)
#as per the data given, x_*1 of 70 inches has brown hair and x_*2 of 68 inches has strawberry hair.
```

We now add an argument `d` representing any legal distance function to the `nn_algorithm_predict` function. Update the implementation so it performs NN using that distance function. Set the default function to be the Euclidean distance in the original function. Also, alter the documentation in the appropriate places.

```{r}
#' Nearest Neighbors Model
#'
#' It classifies its predictions using its nearest neighbor in input space. 
#'
#' @param Xinput      Historical measurements
#' @param y_binary    Historical responses
#' @param Xtest       A matrix of measurements where each row is a unit you wish to predict the response for
#' @param d           Distance function that measures distances b/w two observations, which defaults to squared Euclidean distances.
#' @return            A vector of predicted responses
nn_algorithm_predict = function(Xinput, y_binary, Xtest, d=function(v_1,v_2){
  sum((v_1 - v_2)^2)
}
)
  {
  y_hat=array(NA,nrow(Xtest))
  for (i_star in 1:nrow(Xtest)){
    #y_binary[which.min((Xinput - Xtest[i,])^2)]
    dsq= array(NA,nrow(Xinput))
    for (i in 1:nrow(Xinput)){
      dsq[i]=d(Xinput[i,],Xtest[i,])
    }
    y_hat[i_star]=y_binary[which.min(dsq)]
  }
  y_hat
}
```

For extra credit (unless you're a masters student), add an argument `k` to the `nn_algorithm_predict` function and update the implementation so it performs KNN. In the case of a tie, choose yhat randomly. Set the default `k` to be the square root of the size of D which is an empirical rule-of-thumb popularized by the "Pattern Classification" book by Duda, Hart and Stork (2007). Also, alter the documentation in the appropriate places.

```{r}
#TO-DO for the 643 students but extra credit for undergrads
#need sample mode function from class notes
#' Nearest K Neighbors Model
#'
#' It classifies its predictions using its nearest neighbor in input space. 
#'
#' @param Xinput      Historical measurements
#' @param y_binary    Historical responses
#' @param Xtest       A matrix of measurements where each row is a unit you wish to predict the response for
#' @param d           Distance function that measures distances b/w two observations, which defaults to squared Euclidean distances.
#' @return            A vector of predicted responses
knn_algorithm_predict = function(Xinput, y_binary, Xtest, d=function(v_1,v_2){
  sum((v_1 - v_2)^2)
}
)
  {
  y_hat=array(NA,nrow(Xtest))
  for (i_star in 1:nrow(Xtest)){
    #y_binary[which.min((Xinput - Xtest[i,])^2)]
    dsq= array(NA,nrow(Xinput))
    for (i in 1:nrow(Xinput)){
      dsq[i]=d(Xinput[i,],Xtest[i,])
    }
    y_hat[i_star]=y_binary[which.min(dsq)]
  }
  y_hat
}
```

## Regression via OLS with one feature

Let's quickly recreate the sample data set from practice lecture 7:

```{r}
set.seed(1984)
n = 20
x = runif(n)
beta_0 = 3
beta_1 = -2
```

Compute h^* as `h_star_x`, then draw epsilon from an iid N(0, 0.33^2) distribution as `epsilon`, then compute the vector y.

```{r}
h_star_x = beta_0 + (beta_1*x)
epsilon = rnorm(n, mean =0, sd = .33)
y = h_star_x + epsilon
```

Graph the data by running the following chunk:

```{r}
pacman::p_load(ggplot2)
simple_df = data.frame(x = x, y = y)
simple_viz_obj = ggplot(simple_df, aes(x, y)) + 
  geom_point(size = 2)
simple_viz_obj
```

Does this make sense given the values of beta_0 and beta_1?

A: Yes. 

Write a function `my_simple_ols` that takes in a vector `x` and vector `y` and returns a list that contains the `b_0` (intercept), `b_1` (slope), `yhat` (the predictions), `e` (the residuals), `SSE`, `SST`, `MSE`, `RMSE` and `Rsq` (for the R-squared metric). Internally, you can only use the functions `sum` and `length` and other basic arithmetic operations. You should throw errors if the inputs are non-numeric or not the same length. You should also name the class of the return value `my_simple_ols_obj` by using the `class` function as a setter. No need to create ROxygen documentation here.

```{r}
my_simple_ols = function(x, y){
  ols_obj = list()
  if(class(x) != "numeric"){
    stop("x needs to be numeric")
  }
  
  if(class(x) != "numeric"){
    stop("y needs to be numeric")
  }
  n = length(x)
  
  if (length(y) != n){
    stop("x & y aren't of the same length")
  }
  x_bar = mean(x)
  y_bar = mean(y)
  b_1=(sum(x*y)-(n*x_bar*y_bar))/(sum(x^2)-(n*x_bar^2))
  b_0=y_bar-b_1*x_bar
  yhat = b_0+b_1*x
  e = y - (b_0+b_1*x)
  SSE = sum(e^2)
  MSE = SSE/(n-2)
  RMSE = sqrt(MSE)
  SST = sum((y - y_bar)^2)
  Rsq=((SST-SSE)/SST)
  
  class(ols_obj) = "my_simple_ols_obj"
  ols_obj$b_1 = b_1
  ols_obj$b_0 = b_0
  ols_obj$yhat = yhat
  ols_obj$e = e
  ols_obj$SSE = SSE
  ols_obj$MSE = MSE
  ols_obj$RMSE = RMSE
  ols_obj$SST =  SST
  ols_obj$Rsq = Rsq
  
  ols_obj
}
```

Verify your computations are correct for the vectors `x` and `y` from the first chunk using the `lm` function in R:

```{r}
lm_mod = lm(y ~ x)
my_simple_ols_mod = my_simple_ols(x,y)
#run the tests to ensure the function is up to spec
pacman::p_load(testthat)
expect_equal(my_simple_ols_mod$b_0, as.numeric(coef(lm_mod)[1]), tol = 1e-4)
expect_equal(my_simple_ols_mod$b_1, as.numeric(coef(lm_mod)[2]), tol = 1e-4)
expect_equal(my_simple_ols_mod$RMSE, summary(lm_mod)$sigma, tol = 1e-4)
expect_equal(my_simple_ols_mod$Rsq, summary(lm_mod)$r.squared, tol = 1e-4)
```

Verify that the average of the residuals is 0 using the `expect_equal`. Hint: use the syntax above.

```{r}
expect_equal(0,mean(my_simple_ols_mod$e),tol = 1e-4)
```

Create the $X$ matrix for this data example. Make sure it has the correct dimension.

```{r}
X = cbind(1,x)
```

Use the `model.matrix` function to compute the matrix `X` and verify it is the same as your manual construction.

```{r}
Xstar = data.frame(x)
X_ = model.matrix(~ x, Xstar)
identical(dim(X_),dim(X))
```

Create a prediction method `g` that takes in a vector `x_star` and `my_simple_ols_obj`, an object of type `my_simple_ols_obj` and predicts y values for each entry in `x_star`. 

```{r}
g = function(my_simple_ols_obj, x_star){
  yvalues= my_simple_ols_obj$b_0 + ((my_simple_ols_obj$b_1)*(x_star))
  
  yvalues
}
```

Use this function to verify that when predicting for the average x, you get the average y.

```{r}
expect_equal(g(my_simple_ols_obj, mean(x)), mean(y))
```

In class we spoke about error due to ignorance, misspecification error and estimation error. Show that as n grows, estimation error shrinks. Let us define an error metric that is the difference between b_0 and b_1 and beta_0 and beta_1. How about ||b - beta||^2 where the quantities are now the vectors of size two. Show as n increases, this shrinks. 

```{r}
beta_0 = 3
beta_1 = -2
beta = c(beta_0, beta_1)

ns = c(4,5,25,125,625,4324,100000)
errors = array(NA, length(ns))
for (i in 1 : length(ns)) {
  n = ns[i]
  x = runif(n)
  h_star_x = beta_0 + beta_1 * x 
  epsilon = rnorm(n, mean = 0, sd = 0.33)
  y = h_star_x + epsilon
  
  b=my_simple_ols(x,y)
  errors[i] = sum((beta[[1]] - b$b_0)^2 + (beta[[2]] - b$b_1)^2)
}
errors
```


We are now going to repeat one of the first linear model building exercises in history --- that of Sir Francis Galton in 1886. First load up package `HistData`.

```{r}
pacman::p_load(HistData)
```

In it, there is a dataset called `Galton`. Load it up.

```{r}
data(Galton)
```

You now should have a data frame in your workspace called `Galton`. Summarize this data frame and write a few sentences about what you see. Make sure you report n, p and a bit about what the columns represent and how the data was measured. See the help file `?Galton`.

p is 1 and n is 928 the number of observations. Input is the average parent height and Output is the average child height. 

```{r}
pacman::p_load(skimr)
skim(Galton)
summary(Galton)
```

Find the average height (include both parents and children in this computation).

```{r}
avg_height = mean(c(Galton$parent, Galton$child))
```

If you were predicting child height from parent and you were using the null model, what would the RMSE be of this model be?

```{r}
rmse=sd(Galton$child)
```

Note that in Math 241 you learned that the sample average is an estimate of the "mean", the population expected value of height. We will call the average the "mean" going forward since it is probably correct to the nearest tenth of an inch with this amount of data.

Run a linear model attempting to explain the childrens' height using the parents' height. Use `lm` and use the R formula notation. Compute and report b_0, b_1, RMSE and R^2. 

```{r}
lm_mod = lm(child ~ parent, data=Galton)
coefs = coef(lm_mod)
b_0 = coefs[1]
b_1 = coefs[2]
summary(lm_mod)$r.square
```

Interpret all four quantities: b_0, b_1, RMSE and R^2. Use the correct units of these metrics in your answer.

b_0 = 23.9 is the height of the shortest child
b_1 =0.646<1 which shows average height of children increases at a slower rate than their parents'
rmse = 2.51 inches error, 
r^2 = 0.21 

How good is this model? How well does it predict? Discuss.

The model does not seem to be good as R-squared is 21%, and the rmse is too high.

It is reasonable to assume that parents and their children have the same height? Explain why this is reasonable using basic biology and common sense.

It is reasonable to assume that parents and their children have the same height. Children get their genetics from their parents, therefore one would expect their physical characteristics are the same. 

If they were to have the same height and any differences were just random noise with expectation 0, what would the values of beta_0 and beta_1 be?

beta_0 would be 0 and beta_1 would be 1. 

Let's plot (a) the data in D as black dots, (b) your least squares line defined by b_0 and b_1 in blue, (c) the theoretical line beta_0 and beta_1 if the parent-child height equality held in red and (d) the mean height in green.

```{r}
pacman::p_load(ggplot2)
ggplot(Galton, aes(x = parent, y = child)) + 
  geom_point() + 
  geom_jitter() +
  geom_abline(intercept = b_0, slope = b_1, color = "blue", size = 1) +
  geom_abline(intercept = 0, slope = 1, color = "red", size = 1) +
  geom_abline(intercept = avg_height, slope = 0, color = "darkgreen", size = 1) +
  xlim(63.5, 72.5) + 
  ylim(63.5, 72.5) +
  coord_equal(ratio = 1)
```

Fill in the following sentence: 

Children of short parents became taller on average and children of tall parents became shorter on average.

Why did Galton call it "Regression towards mediocrity in hereditary stature" which was later shortened to "regression to the mean"?

Galton called it such because heights 'regressed' towards the ordinary (synonym for mediocre).

Why should this effect be real?

We are told that average heights are distributed normally. Therefore, if taller parents kept on having tall/taller kids, it would shift the average heights over time. 

You now have unlocked the mystery. Why is it that when modeling with y continuous, everyone calls it "regression"? Write a better, more descriptive and appropriate name for building predictive models with y continuous.

Gaussianization, as the future heights tend to go towards the mean over time. 

You can now clear the workspace.

```{r}
rm(list = ls())
```

Create a dataset D which we call `Xy` such that the linear model has R^2 about 50\% and RMSE approximately 1.

```{r}
x = c(1,2,3)
y = sin(x^2)+x*1.015
Xy = data.frame(x = x, y = y)
my_simple_ols(x,y)$Rsq; my_simple_ols(x,y)$RMSE
```

Extra credit but required for 642 students: create a dataset D and a model that can give you R^2 arbitrarily close to 1 i.e. approximately 1 - epsilon but RMSE arbitrarily high i.e. approximately M.

```{r}
epsilon = 0.01
M = 1000
x = c(1,2,10)
y = c(200,800,17800)
#y = c(200,1200,6000)
Xy = data.frame(x = x, y = y)
my_simple_ols(x,y)$Rsq; my_simple_ols(x,y)$RMSE
```






